\usefont{OT1}{bch}{b}{n} \normalfont \normalsize \textsc{STEM Fellowship Big Data Challenge 2020} \\ [10pt] \huge Deep learning transcriptomic model for prediction of pan-drug chemotherapeutic sensitivity \\

### 2020-07-05 manuscript/manuscript.tex
updated manuscript

\begin{document}
% Abstract
\twocolumn[
	\maketitle
	\begin{abstract}
		The emergence of precision oncology approaches have begun to inform clinical decision-making in diagnostic, prognostic, and treatment contexts. High-throughput technology has enabled machine learning algorithms to utilize molecular characteristics of tumours to generate personalized treatments, revealing relationships between genomic alterations and drug sensitivity. However, precision oncology studies have yet to generate a predictive biomarker incorporating gene expression profiles to stratify tumours into similar pan-drug sensitivity profiles. Here we show that a neural network with 10 hidden layers accurately classifies pan-cancer cell lines into two distinct chemotherapeutic response groups with respect to a pan-drug dataset with 89.0\% accuracy (AUC = 0.904, weighted F1 score = 0.89). From unsupervised clustering algorithms, we found that a cohort of cell line gene expression data from the Genomics of Drug Sensitivity in Cancer could be clustered into two response groups with significant differences in pan-drug chemotherapeutic sensitivity. We used the Boruta feature selection algorithm to identify genes of the highest predictive value; the 300 selected genes were enriched primarily for the focal adhesion, ECM-receptor and proteoglycan interaction pathways. Using the selected genes, a deep learning neural network was developed to predict response groups. The classification efficacy of the neural network validates our postulate that cell lines with similar gene expression profiles present similar pan-drug chemotherapeutic sensitivity, and it provides evidence for the potential utility of similar combinatorial biomarkers for the selection of potent candidate drugs. Given that not all tumours present with targetable features and are treated with less targeted and consequently more cytotoxic chemotherapies, our computational framework allows for the identification of candidate drugs for tailored treatments that maximize therapeutic response and minimize cytotoxic burden.
		\vspace{1em}

@@

Here we set out to generate a deep learning transcriptomic model for the prediction of pan-drug chemotherapeutic sensitivity across cell lines of all cancer types. If successful, this would demonstrate that gene expression influences chemotherapeutic response across most drugs and further motivate future studies into the development of clinically applicable predictors of candidate chemotherapeutics for tumours of a specific gene expression profile. Further, this biomarker may prove to be a useful technique to stratify patients into distinct treatment response groups, allowing chemotherapy regimens to be tailored to the transcriptomic characteristics of a particular cancer.

@@

\subsection{Determining pan-cancer therapeutic response cohorts}
To better understand the impact and predictive ability of transcriptomic dysregulation in chemotherapeutic response, a pan-cancer cohort of cell lines and associated therapeutic sensitivity data were obtained from the GDSC database. The database includes 1,110 cell lines from various tumour types and is thought to represent a relatively comprehensive pan-cancer dataset. In addition, the acquired dataset contained therapeutic efficacy information in the form of half-maximal inhibitory concentration (IC$_{50}$) values for 251 chemotherapies. These values correspond to the minimal concentration of therapeutic required to induce cell death in 50\% of the cultured cells, with lower values being associated with improved drug efficacy. The dataset was used to generate a matrix with cell lines and accompanying therapeutic information. The dataset was then filtered to exclude therapies with less than 80\% of data for all cell lines, followed by the exclusion of cell lines lacking response data for the drugs retained in the first step. The filtered dataset was analyzed to ensure that it was still representative of a relatively pan-cancer dataset.

@@

Following the evaluation of existing classifiers, we attempted to create defined cell line clusters using observed chemotherapeutic response of the pan-cancer cell line samples. We developed a normalized Euclidean distance matrix for the retained cell lines based on their pan-chemotherapy response. The matrix was used to identify the minimum number of clusters capable of representing the therapeutic heterogeneity identified across the cancer cell lines while maintaining significant inter-cluster distance. From the matrix, k-means clustering was utilized to assign cell line candidates to appropriate therapeutic response cohorts (the elbow method was used as a heuristic for determining the number of clusters). Generalized differences in chemotherapeutic efficacy between cohorts were visualized using a heatmap generated by the pheatmap package \cite{pheatmap} in R. The heatmap displayed 2 groups clustered using k-means clustering. The separation between clusters was also visualized using principal component analysis with the plotly package \cite{plotly} in R. Following the identification of defined clusters, differences in therapeutic efficacy between the identified cohorts were evaluated. Mann-Whitney U tests were utilized to compare IC$_{50}$ values between the groups. False discovery rate (FDR) correction was utilized to correct for multiple comparisons. To further validate the existence of these distinct chemotherapeutic response  clusters, intra-cancer heterogeneity was evaluated. 11 cancer subtypes with sufficient cell line representation ($n>20$) were identified. Differences in chemotherapeutic efficacy between cell lines assigned to group A and B were visualized using heatmaps.

@@

Following grid search and cross-validation, we evaluated the performance of our model on the testing data. Youden's index from the ROC curves corresponding to each neural network’s performance on the training dataset was used to determine the optimal threshold to classify cell lines into chemotherapeutic response groups for the testing dataset. Youden’s index ranges from 0 to 1 inclusive, and an index of 0 indicates that the binary classifier is no better than that obtained from a fair coin flip. Conversely, a value of 1 indicates that the test is perfect (i.e., no false positives or false negatives). The formula for Youden's index is defined as $$J \doteq \text{sensitivity} + \text{specificity} - 1$$ All indices were obtained from the ROC curves by an iterative search for the maximum $J$. That is, $$J = \text{argmax(TPR} - \text{FPR)}$$ where TPR and FPR ($\text{FPR} = 1 - \text{specificity}$) are arrays containing the true positive and false positive rate corresponding to each point on the ROC curve. The threshold corresponding to $J$ was used to classify cell lines.

>>>>

\begin{document}
% Abstract
\twocolumn[
	\maketitle
	\begin{abstract}
		The emergence of precision oncology approaches have begun to inform clinical decision-making in diagnostic, prognostic, and treatment contexts. High-throughput technology has enabled machine learning algorithms to utilize molecular characteristics of tumours to generate personalized treatments, revealing relationships between genomic alterations and drug sensitivity. However, precision oncology studies have yet to generate a predictive biomarker incorporating gene expression profiles to stratify tumours into similar pan-drug sensitivity profiles. Here we show that a neural network with 10 hidden layers accurately classifies pan-cancer cell lines into two distinct chemotherapeutic response groups with respect to a pan-drug dataset with 89.0\% accuracy (AUC = 0.904, weighted F1 score = 0.89). From unsupervised clustering algorithms, we found that a cohort of cell line gene expression data from the Genomics of Drug Sensitivity in Cancer could be clustered into two response groups with significant differences in pan-drug chemotherapeutic sensitivity. We used the Boruta feature selection algorithm to identify genes of the highest predictive value; the 300 selected genes were enriched primarily for the focal adhesion, ECM-receptor and proteoglycan interaction pathways. Using the selected genes, a deep learning neural network was developed to predict response groups. The high classification efficacy of the neural network validates our postulate that cell lines with similar gene expression profiles present similar pan-drug chemotherapeutic sensitivity, and it provides evidence for the potential utility of similar combinatorial biomarkers for the selection of potent candidate drugs. Given that not all tumours present with targetable features and are treated with less targeted and consequently more cytotoxic chemotherapies, our computational framework allows for the identification of candidate drugs for tailored treatments that maximize therapeutic response and minimize cytotoxic burden.
		\vspace{1em}

@@

In this study, we set out to generate a deep learning transcriptomic model for the prediction of pan-drug chemotherapeutic sensitivity across cell lines of all cancer types. If successful, this would demonstrate that gene expression influences chemotherapeutic response across most drugs and further motivate future studies into the development of clinically applicable predictors of candidate chemotherapeutics for tumours of a specific gene expression profile. Further, this biomarker may prove to be a useful technique to stratify patients into distinct treatment response groups, allowing chemotherapy regimens to be tailored to the transcriptomic characteristics of a particular cancer.

@@

\subsection{Determining pan-cancer therapeutic response cohorts}
To better understand the impact and predictive ability of transcriptomic dysregulation in chemotherapeutic response, a pan-cancer cohort of cell lines and associated therapeutic sensitivity data were obtained from the GDSC database. The database includes 1,110 cell lines from various tumour types and is thought to represent a relatively comprehensive pan-cancer dataset. In addition, the acquired dataset contained therapeutic efficacy information in the form of half-maximal inhibitory concentration (IC$_{50}$) values for 251 chemotherapies. These values correspond to the minimal concentration of therapeutic required to induce cell death in 50\% of the cultured cells, with lower values being associated with improved drug efficacy. The dataset was used to generate a matrix with cell lines and accompanying therapeutic information. The matrix was then filtered to exclude therapies with less than 80\% of data for all cell lines, followed by the exclusion of cell lines lacking response data for the drugs retained in the first step. The filtered matrix was analyzed to ensure that it was still representative of a relatively pan-cancer dataset.

@@

Following the evaluation of existing classifiers, we attempted to create defined cell line clusters using observed chemotherapeutic response of the pan-cancer cell line samples. We developed a normalized Euclidean distance matrix for the retained cell lines based on their pan-chemotherapy response. The matrix was used to identify the minimum number of clusters capable of representing the therapeutic heterogeneity identified across the cancer cell lines while maintaining significant inter-cluster distance. From the matrix, k-means clustering was utilized to assign cell line candidates to appropriate therapeutic response cohorts (the elbow method was used as a heuristic for determining the number of clusters). Generalized differences in chemotherapeutic efficacy between cohorts were visualized using a heatmap generated by the pheatmap package \cite{pheatmap} in R. The heatmap displayed 2 groups clustered using k-means clustering. The separation between clusters was also visualized using principal component analysis with the plotly package \cite{plotly} in R. Following the identification of defined clusters, differences in therapeutic efficacy between the identified cohorts were evaluated. Mann-Whitney U tests were utilized to compare IC$_{50}$ values between the groups. False discovery rate (FDR) correction was utilized to correct for multiple comparisons. To further validate the existence of these distinct chemotherapeutic response  clusters, intra-cancer heterogeneity was evaluated; 11 cancer subtypes with sufficient cell line representation ($n>20$) were identified. Differences in chemotherapeutic efficacy between cell lines assigned to group A and B were visualized using heatmaps.

@@

Following grid search and cross-validation, we evaluated the performance of our model on the testing data. Youden's index from the ROC curves corresponding to each neural network’s performance on the training dataset was used to determine the optimal threshold to classify cell lines into chemotherapeutic response groups for the testing dataset. Youden’s index ranges from 0 to 1 inclusive, and an index of 0 indicates that the binary classifier is no better than that obtained from a fair coin flip. Conversely, a value of 1 indicates that the test is perfect (i.e., no false positives or false negatives). The formula for Youden's index is defined as $$J \doteq \text{sensitivity} + \text{specificity} - 1$$ All indices were obtained from the ROC curves by an iterative search for the maximum $J$. That is, $$J = \text{argmax(TPR} - \text{FPR)}$$ where TPR and FPR ($\text{FPR} = 1 - \text{specificity}$) are arrays containing the true positive and false positive rate corresponding to each point on the ROC curve. The threshold corresponding to $J$ was used to classify cell lines. The neural network performance on the testing dataset was visualized using ROC curves and confusion matrices.
